A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-10-03 tex/POPL24/sec2-outline.tex
[ tex ] Add pdf

The construction is straightforward:
Run the mode decorator on the input term~$\erase\Gamma \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $\erase\Gamma \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}).

\begin{lemma}[Soundness]
If\/ $\Gamma \vdash t \syn A$, then $\Gamma \vdash t : A$.
\end{lemma}

If there is no derivation of $\Gamma \vdash t \syn A$ for any type~$A$, report that there is no derivation of $\Gamma \vdash t : A$ for any~$A$ either (the second outcome), because the bidirectional typing is \emph{complete} with respect to the original typing.

\begin{lemma}[Completeness] \label{thm:STLC-completeness}
If\/ $\erase\Gamma \vdash t^{\syn}$ and\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
\end{lemma}

>>>>

The construction is straightforward:
Run the mode decorator on the input term~$\erase\Gamma \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $\erase\Gamma \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}); if there is no derivation of $\Gamma \vdash t \syn A$ for any type~$A$, then there is no derivation of $\Gamma \vdash t : A$ for any~$A$ either (the second outcome), because the bidirectional typing is \emph{complete}:
\begin{theorem}[Soundness and Completeness]
$\Gamma \vdash t \syn A$ if and only if\/ $\erase\Gamma \vdash t^{\syn}$ and\/ $\Gamma \vdash t : A$.
\end{theorem}
%
%
%\begin{lemma}[Completeness] \label{thm:STLC-completeness}
%If\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
%\end{lemma}

### 2023-10-03 tex/POPL24/sec7-future.tex
[ tex ] Add pdf

\section{Concluding Remarks} \label{sec:future}

@@

As a first theory of bidirectional type synthesis, our work focuses on syntax-directed type systems.
To explore more general settings, an extreme possibility is to independently specify an ordinary type system and a bidirectional one over the same raw terms and then investigate possible relationships between the two systems, but a more manageable and practical setting suitable for a next step is probably one that only mildly generalises ours: 
for each raw term construct, there can be several ordinary typing rules, and each typing rule can be refined to several bidirectional typing rules with different mode assignments.
Even for this mildly generalised setting, there is already a lot more work to do:
Both the mode decorator and the bidirectional type synthesiser will have to backtrack, and become significantly more complex.
For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, but the completeness direction will pose a problem---for every node of a raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
There should be ways to fix this problem while retaining the mode decoration phase---for example, one idea is to make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of these mode derivations into a bidirectional typing derivation.
All these considerations are rather too complex for a first theory, however, and we decide to leave them for future work.

@@

\paragraph{Beyond simple types}
While algebraic approaches to polymorphic types have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
Subtyping is essential for formulating important concepts such as \emph{principal types} in type synthesis.
On the other hand, in the realm of dependent types, \citeauthor{Cartmell1986}'s generalised algebraic theories~\citeyearpar{Cartmell1986} can handle a wide variety of dependent type theories.
\citet{Bezem2021} investigate the notion of presentation (extending the notion of signature) in the context of generalised algebraic theories.
Nonetheless, type synthesis for dependent types requires normalisation or some form of conversion to check type equality.
Normalisation in its generic form still remains out of reach and recent advances in this topic are discussed in the doctoral thesis by~\citet{Valliappan2023}.

\paragraph{Whither extrinsic typing?}

\todo[inline]{We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}), which requires advancement in language formalisation~(\cref{sec:language-formalisation}) that is orthogonal to our work.}

\Josh[inline]{Optional: ornaments (0.25p), NDGP (0.25p)}

>>>>

\section{Future Work} \label{sec:future}

\paragraph{More characteristics of bidirectional typing}

We have formalised some of the most important concepts discussed by \citet{Dunfield2021}.
There are other concepts that may be formalisable with some more effort, for example Pfenning's recipe for bidirectionalising typing rules.
We believe that our theory lays the foundation for further formalising these concepts.
There are some other concepts that may be more difficult to pin down, notably `annotation character', which is roughly about how easy it is for the user to write annotated programs.
These may continue to require creativity from the designer of type systems.

@@

As a first theory of bidirectional type synthesis, our work focuses on syntax-directed type systems.
To explore more general settings, an extreme possibility is to independently specify an ordinary type system and a bidirectional one over the same raw terms and then investigate possible relationships between the two systems, but a more manageable and practical setting suitable for a next step is probably one that only mildly generalises ours: 
for each raw term construct, there can be several ordinary typing rules, and each typing rule can be refined to several bidirectional typing rules with different mode assignments.
Even for this mildly generalised setting, there is already a lot more work to do:
Both the mode decorator and the bidirectional type synthesiser will have to backtrack, and become significantly more complex.
For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, but the completeness direction will pose a problem---for every node of a raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
There should be ways to fix this problem while retaining the mode decoration phase---for example, one idea is to make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of these mode derivations into a bidirectional typing derivation.
All these considerations are rather too complex for a first theory, however.

@@

\paragraph{Beyond simple types}
We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}).
While algebraic approaches to polymorphic types have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
Subtyping is essential for formulating important concepts such as \emph{principal types} in type synthesis.
On the other hand, in the realm of dependent types, \varcitet{Cartmell1986}{'s} generalised algebraic theories can handle a wide variety of dependent type theories.
\citet{Bezem2021} investigate the notion of presentation (extending the notion of signature) in the context of generalised algebraic theories.
Nonetheless, type synthesis for dependent types requires normalisation or some form of conversion to check type equality.
Normalisation in its generic form still remains out of reach, and advances in this topic are only recently discussed in the doctoral thesis by~\citet{Valliappan2023}.
Before we can embark on a richer theory of bidirectional type synthesis, language formalisation frameworks for these more advanced language features have to be developed.

%\paragraph{Whither extrinsic typing?}

%\todo[inline]{We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}), which requires advancement in language formalisation~(\cref{sec:language-formalisation}) that is orthogonal to our work.}

%\Josh[inline]{Optional: ornaments (0.25p), NDGP (0.25p)}
